---
title: "PlantPhys with Dr. Danny"
author: "Susan Bush, Danny Kaplan"
date: "June 22, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(tidyr)
# library(printr)
```

## Reading the "raw" data

The data have been stored by the FANCY plate reader in a spreadsheet-like file, called CSV format. We're going to read from that file to create an object in R that contains the data.

Note: It's bad practice to do excel kinds of things, like adding up a column in your data. Your data are sacred. Don't muck about with them, unless you want to be doing scientific fraud. In which case, take Psych 308.

```{r}
library(ggplot2)
library(tidyr)
my_data_folder <- "~/Documents/Mac/2015-2016 school year/PlantPhys Fall15/L04.big data lab/"
my_data_file <- "WildSpecies_ColdRoots.csv"

# when you save your data to CSV
# write.csv(dataframe, file = "me.csv", row.names = FALSE)

Roots <- read.csv(paste0(my_data_folder, my_data_file), row.names = 1)
```

We'll clean things up a little bit. Like change the names of the columns.
```{r}
names(Roots) <- gsub("root", "day", names(Roots))
```

# The data look like this

```{r echo = FALSE}
head(Roots, 10)
```

# What have we got?

Compare Day 6 length for the different plants

```{r}
Roots %>%
  ggplot(aes(x = day6)) + geom_density(fill = "gray", color = NA)
```

For the different plates:
```{r}
Roots %>% 
  ggplot(aes(x = day6)) + geom_histogram() + facet_grid( spp ~ trt)
```


# What if we condense (reduce) the data?

```{r}
Summary <-
  Roots %>%
  select(trt, plate, spp, day6) %>%
  group_by(trt, spp) %>%
  summarise(typical = mean(day6, na.rm = TRUE), stdev = sd(day6, na.rm = TRUE) / sqrt(n()))
```

Make a graph
```{r}
Summary %>%
  ggplot(aes(x = spp, y = typical, fill = spp)) + 
  geom_bar(stat = "identity", alpha = .2) + 
  geom_errorbar(aes(ymax = typical + 2 * stdev, ymin = typical - 2 * stdev), width = .1) + 
  geom_point(data = Roots, aes(x = spp, y = day6), alpha = .3) +
  facet_grid( ~ trt)
```